Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-29 ms.tex
going back to old TI prescription.

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to
measure the radial surface brightness profiles for hundreds of nearby
early type galaxies. The measured profile is fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law
slope, $\Gamma$, to an outer power law slope, $\beta$, at a break
radius, $\rb$.  If the stellar population is spherically symmetric,
then this corresponds to a 3D stellar density $\rhostar \propto
r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for
$r \gg \rb$.  We can write the deprojected stellar density
approximately (formally, this is the $\alpha \rightarrow \infty$
limit) as
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rinf} \left(r/\rinf\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rinf}$ is the stellar density at the radius of
the black hole sphere of influence\footnote{This approximate
  expression agrees surprisingly well with the mean empirical scaling
  relation for $r_{\rm inf}(M_\bullet)$ calculated in Appendix C of
  \citet{Stone&Metzger15}, although we note that this relation has
  significant scatter.},
\begin{equation}
  \rinf \simeq G \Mbh/\sigma_{\bullet}^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the
second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$
relationship of \citet{McConnellMa+:2011a},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma_{\bullet}}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g.,
\citealt{Greene+2010, Kormendy+2013}), we adopt equation
(\ref{eq:Msigma}) for lack of an obviously better alternative.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{TI.pdf}
  \caption{Minimum effective wind heating parameter required for
    thermal stability as a function of SMBH mass.  Black lines show
    $v_{\rm TI}$ (eq.~[\ref{eq:cooling3}]), the heating rate required
    for $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}} > 10$ in the
    high-heating limit when the stagnation radius lies interior to the
    influence radius, for different values of the mass loss parameter
    $\eta$ as marked.  Blue lines show the minimum heating
    parameter required to have $\zeta > \zeta_c = (r_{b}/r_{\rm
      inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]), separately for
    cusp (solid) and core (dashed) galaxies, using the mean
    $r_{b}/r_{\rm inf} \approx 7.6M_{\bullet,8}^{-0.2}$ from the
    \citet{LauerFaber+:2007a} galaxy sample for cores and assuming a fixed value of $r_{b} = 100$ pc for cusps.  For $\zeta < \zeta_c$
    the stagnation radius moves from inside the influence radius, out
    to the stellar break radius $r_b$ (Appendix \ref{app:be}); in
    practice renders the flow susceptible to thermal runaway, even if
    $v_{w} > v_{\rm TI}$.  For cusp galaxies the $v_w > v_{\rm TI}$ is
    usually the more strigent criterion for stability, while for core
    galaxies the $\zeta > \zeta_c$ is usually more relevant.  }
\label{fig:TI}
\end{figure}

@@

Based on our numerical results (\S \ref{sec:numerical}) and the work of \citet{McCourt+12} we define thermally stable flows according to the criterion $\dot{q}_{\rm heat} > 10|\dot{q}_{\rm rad}|$ ($t_{\rm
  cool} > 10t_{\rm ff}$) being satisfied near the stagnation radius.  This condition translates into a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  \begin{cases}
   280 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   240 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}  \\
%\approx \sigma_0(r_{b}/r_{\rm inf})^{0.5(1-\Gamma)} &, \zeta \ll \zeta_c. 
  \end{cases}
  \label{eq:cooling3}
\end{align}
Equations (\ref{eq:cooling2}) and (\ref{eq:cooling3}) are derived using expressions for the stagnation radius and gas density in the high heating limit of $\zeta > \zeta_c = (r_{b}/r_{\rm inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]).  However, for $\zeta < \zeta_c$, the stagnation radius diverges to the break radius $r_b$ (Appendix \ref{app:be}).  The quasi-hydrostatic structure that results in this case greatly increases the gas density, which in practice renders the flow susceptible to thermal runaway, even if $v_{w} > v_{\rm TI}$ according to equation (\ref{eq:cooling3}).  In other words, the true condition for thermal instability  can be written
\be
v_w > {\rm max}[v_{\rm TI},\sigma_0\{1-(r_{b}/r_{\rm inf})^{0.5(1-\Gamma)}\}]
\ee

Figure \ref{fig:TI} shows $v_{\rm TI}(M_{\bullet})$ for different mass input parameters $\eta$, as well as the [$\eta$-independent] $\zeta > \zeta_c$ criterion, shown separately for cusp and core galaxies.  For core galaxies, we calculate $\zeta_c$ using the empirical relation, $r_{b}/r_{\rm inf} \approx 7.6M_{\bullet,8}^{-0.2}$, which
follows from equation (\ref{eq:rsoi}) and the $r_b-M_{\bullet}$
correlation derived from the results of \citet{LauerFaber+:2007a}.
For cusp galaxies we see that the $v_w \gtrsim v_{\rm TI}$ criterion is generally more strigent, while for core galaxies the $\zeta > \zeta_c$ criterion is more strigent for core galaxies, except for high mass loss rates $\eta > 1$.

The minimum heating rate for thermal stability (eq.~[\ref{eq:cooling2}]) corresponds to {\it maximum} thermally-stable accretion rate,
\begin{align}
  \frac{\dot{M}_{\rm TI}}{\dot{M}_{\rm edd}} \simeq \begin{cases}
    4\times 10^{-6}\eta_{0.02}M_{\bullet,8}^{-0.37} &, \text{core}, \\
    8\times 10^{-5} \eta_{0.02}^{0.59} M_{\bullet,8}^{0.28}    &, \text{cusp},\\
  \end{cases}
  \label{eq:Mdotmax}
\end{align}
where the cusp case is derived using equation (\ref{eq:eddr_analytic}) to calculate $\dot{M}(v_{\rm TI})$, and the core case corresponds to the more stringent $\zeta > \zeta_c$ criterion, calculated again using the mean $r_{b}/r_{\rm inf}$ relation from \citet{LauerFaber+:2007a}.

>>>>

\section{Model}
\label{sec:model}
\subsection{Galaxy models}
\label{sec:gal_model}
\citet{LauerFaber+:2007a} use Hubble Space Telescope WFPC2 imaging to
measure the radial surface brightness profiles for hundreds of nearby
early type galaxies. The measured profile is fit by a ``Nuker" law parameterization:
\begin{equation}
  I(\xi)=I_b 2^{(\beta-\Gamma)/\alpha} \xi^{-\Gamma} (1+\xi^\alpha)^{-(\beta-\Gamma)/\alpha}, \,\,\,\xi\equiv\frac{r}{r_b},
\end{equation}
i.e., a broken power law that transitions from an inner power law
slope, $\Gamma$, to an outer power law slope, $\beta$, at a break
radius, $\rb$.  If the stellar population is spherically symmetric,
then this corresponds to a 3D stellar density $\rhostar \propto
r^{-1-\Gamma}$ for $r \ll \rb$ and $\rhostar\propto r^{-1-\beta}$ for
$r \gg \rb$.  We can write the deprojected stellar density
approximately (formally, this is the $\alpha \rightarrow \infty$
limit) as
\begin{align}
\rho_\star = 
\begin{cases}
\rho_\star|_{\rinf} \left(r/\rinf\right)^{-1-\Gamma} & r \leq r_b\\
\rho_\star|_{r_b} \left(r/r_b\right)^{-1-\beta} & r > r_b,
\label{eq:rhostar}
\end{cases}
\end{align}
where $\rho_\star|_{\rinf}$ is the stellar density at the radius of
the black hole sphere of influence\footnote{This approximate
  expression agrees surprisingly well with the mean empirical scaling
  relation for $r_{\rm inf}(M_\bullet)$ calculated in Appendix C of
  \citet{Stone&Metzger15}, although we note that this relation has
  significant scatter. Also, we note that scaling is somewhat
  different for cores and cusps. Fixing the power law slope,
  $\rinf\simeq 25 (8) \Mbheight^{0.6}$ for core (cusp)
  galaxies. Unless otherwise notes we will use the single relation in
  equation~\eqref{eq:rsoi}},
\begin{equation}
  \rinf \simeq G \Mbh/\sigma_{\bullet}^2 \approx 14 M_{\bullet,8}^{0.6}\,{\rm pc},
\label{eq:rsoi}
\end{equation}
where $M_{\bullet,8} \equiv M_{\bullet}/10^{8}M_{\odot}$ and the
second equality in (\ref{eq:rsoi}) employs the $\Mbh-\sigma$
relationship of \citet{McConnellMa+:2011a},
 \begin{align}
M_{\bullet} \simeq 2\times 10^{8}\left(\frac{\sigma_{\bullet}}{200\,{\rm
      km\,s^{-1}}}\right)^{5.1}M_{\odot}.
\label{eq:Msigma}
\end{align}
Though of questionable validity for low mass black holes (e.g.,
\citealt{Greene+2010, Kormendy+2013}), we adopt equation
(\ref{eq:Msigma}) for lack of an obviously better alternative.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{TI.pdf}
  \caption{Minimum effective wind heating parameter required for
    thermal stability as a function of SMBH mass.  Black lines show
    $v_{\rm TI}$ (eq.~[\ref{eq:cooling3}]), the heating rate required
    for $(\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|)_{r_{s}} > 10$ in the
    high-heating limit when the stagnation radius lies interior to the
    influence radius, for different values of the mass loss parameter
    $\eta$ as marked.  Blue lines show the minimum heating parameter
    required to have $\zeta > \zeta_c = (r_{b}/r_{\rm
      inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]), separately for
    cusp (solid) and core (dashed) galaxies. Based on the
    \citet{LauerFaber+:2007a} sample we take $\rinf=25 (8) \Mbh^{0.6}
    pc$ and $\rb=106 \Mbheight^{0.4}$ (100 pc)  for cores (cusps). For
    $\zeta < \zeta_c$ the stagnation radius moves from inside the
    influence radius, out to the stellar break radius $r_b$ (Appendix
    \ref{app:be}); in practice renders the flow susceptible to thermal
    runaway, even if $v_{w} > v_{\rm TI}$.  For cusp galaxies the $v_w
    > v_{\rm TI}$ is usually the more strigent criterion for
    stability, while for core galaxies the $\zeta > \zeta_c$ is
    usually more relevant.}
\label{fig:TI}
\end{figure}

@@

Based on our numerical results (\S \ref{sec:numerical}) and the work of \citet{McCourt+12} we define thermally stable flows according to the criterion $\dot{q}_{\rm heat} > 10|\dot{q}_{\rm rad}|$ ($t_{\rm
  cool} > 10t_{\rm ff}$) being satisfied near the stagnation radius.  This condition translates into a
critical minimum heating rate
\begin{align}
v_{w} > v_{\rm TI} \simeq
  \begin{cases}
   280 \eta_{0.02}^{0.14} M_{\bullet,8}^{0.11}\,{\rm km\,s^{-1}}  &, \text{core}\\
   240 \eta_{0.02}^{0.17} M_{\bullet,8}^{0.08}\,{\rm km\,s^{-1}}   &, \text{cusp}  \\
%\approx \sigma_0(r_{b}/r_{\rm inf})^{0.5(1-\Gamma)} &, \zeta \ll \zeta_c. 
  \end{cases}
  \label{eq:cooling3}
\end{align}
Equations (\ref{eq:cooling2}) and (\ref{eq:cooling3}) are derived
using expressions for the stagnation radius and gas density in the
high heating limit of $\zeta > \zeta_c = (r_{b}/r_{\rm
  inf})^{0.5(1-\Gamma)}$ (eq.~[\ref{eq:zetacrit}]).  However, for
$\zeta < \zeta_c$, the stagnation radius diverges to the break radius
$r_b$ (Appendix \ref{app:be}).  The quasi-hydrostatic structure that
results in this case greatly increases the gas density, which in
practice renders the flow susceptible to thermal runaway, even if
$v_{w} > v_{\rm TI}$ according to equation (\ref{eq:cooling3}).  In
other words, the true condition for thermal instability  can be
written

\begin{equation}
v_w > {\rm max}[v_{\rm TI},\sigma_0\{1-(r_{b}/r_{\rm
  inf})^{0.5(1-\Gamma)}\}]
\label{eq:cooling4}
\end{equation}

Figure \ref{fig:TI} shows $v_{\rm TI}(M_{\bullet})$ for different mass
input parameters $\eta$, as well as the [$\eta$-independent] $\zeta >
\zeta_c$ criterion, shown separately for cusp and core galaxies.
Based on the \citet{LauerFaber+:2007a} sample we take $\rinf = 25(8)M
0.6$ pc and $\rb = 106 \Mbheight^{0.4}$ (100 pc) and thus
$\rb/\rinf\simeq 4 \Mbheight^{-0.2} (13 \Mbheight^{-0.6})$ for cores
(cusps).  For cusp galaxies we see that the $v_w \gtrsim v_{\rm TI}$
criterion is generally more stringent, while for core galaxies the
$\zeta > \zeta_c$ criterion is more strigent for core galaxies, except
for high mass loss rates $\eta > 1$.

The minimum heating rate for thermal stability corresponds to {\it
  maximum} thermally-stable accretion rate. Due to the large scatter
in galaxy structural parameters at a given $\Mbh$, we use
equation~\eqref{eq:cooling3} instead of equation~\eqref{eq:cooling4}:
the $\zeta_c$ criterion should be evaluated on a galaxy-by-galaxy
basis. Thus, the maximum thermally stable $\Mdot$ (for
$\zeta>\zeta_c$) is

\begin{align}
  \frac{\dot{M}_{\rm TI}}{\dot{M}_{\rm edd}} \simeq \begin{cases}
    2\times 10^{-4}\eta_{0.02}^{0.47}M_{\bullet,8}^{0.36} &, \text{core}, \\
    8\times 10^{-5} \eta_{0.02}^{0.59} M_{\bullet,8}^{0.28}    &, \text{cusp},\\
  \end{cases}
  \label{eq:Mdotmax}
\end{align}
where the cusp case is derived using equation (\ref{eq:eddr_analytic})
to calculate $\dot{M}(v_{\rm TI})$, and the core case corresponds to
the more stringent $\zeta > \zeta_c$ criterion, calculated again using
the mean $r_{b}/r_{\rm inf}$ relation from \citet{LauerFaber+:2007a}.
